---
title: "Homework 2"
author: "Jake Mc Leroth"
date: "2025-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Question 2:
```{r data}
data <- read.csv("/Users/jakemcleroth/Desktop/University/Masters/Modules/Semester 1/SLT/homework/hw_2/prostate.csv", header = FALSE)
colnames(data) <- c("lcavol", "lweight","age", "lbph", "svi", "lcp", "gleason", "pgg45", "lpsa", "train")
data <- data[, c("lpsa", "lcavol", "lweight", "age", "lbph", "svi", "lcp", "gleason", "pgg45", "train")]
```


a) Here we construct the scatter plot in Figure 1.1:
```{r a}
pairs(data[,-10],
      pch = 1,
      col = "purple",
      cex = 0.5,
      lwd = 1,
      gap = 0.5,
      main = "Figure 1.1"
)
```

b) We repeat example 3.2.1. First we get the correlation matrix of predictors as table 3.1:
```{r b correlation matrix}
row_vars <- c("lweight", "age", "lbph", "svi", "lcp", "gleason", "pgg45")
col_vars <- c("lcavol", "lweight", "age", "lbph", "svi", "lcp", "gleason")
cor_matrix <- cor(data[data[,10] == TRUE, ][, row_vars], data[data[,10] == TRUE,][, col_vars], use = "complete.obs") # only using training data here
print(round(cor_matrix, 3))
```
Looking at the lower triangel, we see that the values mathc those in table 3.1. We now fit a linear regression model, first splitting scaling the predictor matrix:
```{r data}
data <- read.csv("/Users/jakemcleroth/Desktop/University/Masters/Modules/Semester 1/SLT/homework/hw_2/prostate.csv", header = FALSE)
colnames(data) <- c("lcavol", "lweight","age", "lbph", "svi", "lcp", "gleason", "pgg45", "lpsa", "train")
data <- data[, c("lpsa", "lcavol", "lweight", "age", "lbph", "svi", "lcp", "gleason", "pgg45", "train")]
```

```{r scale}
scaled_data <- data
means <- apply((data[,-10]),2,mean)
sds <- apply(data[, -10], 2, sd)

for (i in 2:9) {
  scaled_data[,i] <- (data[,i] - means[i])/sds[i]
}
```
We then split the data:
```{r training and test}
std_train_data <- scaled_data[data[,10] == TRUE,][,-10]
std_test_data <- scaled_data[data[,10] == FALSE,][,-10]
```

Here we fit the linear regression and show the parameter values
```{r b linear model fit}
model_1 <- lm(lpsa ~ lcavol + lweight + age + lbph + svi + lcp + gleason + pgg45, data = std_train_data)
model_summary_1 <- summary(model_1)
results <- data.frame(
  Coefficient = model_summary_1$coefficients[, "Estimate"],
  Std_Error = model_summary_1$coefficients[, "Std. Error"],
  Z_Score = model_summary_1$coefficients[, "t value"],
  P_Value = model_summary_1$coefficients[, "Pr(>|t|)"]
)

print(round(results,3), row.names = FALSE)
```
We see these are identical to Table 3.2.

c) To obtain the F statistic on pg 51, we must refit the model without the insignficant parameters (age, lcp, gleason and pgg45)

```{r F-statistic}
reduced_model <- lm(lpsa ~ lcavol + lweight + lbph + svi, data = std_train_data)

test_result <- anova(reduced_model, model_1)
f_statistic <- test_result$F[2]
p_value <- test_result$`Pr(>F)`[2]
df_num <- test_result$Df[2]  
df_den <- df.residual(model_1) 

exclusion_test <- data.frame(
  Test = "Partial F-test",
  F_statistic = f_statistic,
  df_num = df_num,
  df_den = df_den,
  p_value = p_value
)

print(exclusion_test)
```
We see here this matches the value and p-value on pg51. Here we show the absolute mean error:

```{r mean error rate}
y_pred <- predict(model_1, newdata = std_test_data)
y_test <- std_test_data$lpsa

prediction_errors <- (y_test - y_pred)^2
mean_prediction_error <- mean(prediction_errors)
print(mean_prediction_error)
```
Which is the same as on page 51.

d) Here, additionally do the Ridge, Lasso and best subset selection
```{r}
library(glmnet)
```

Column 4 of table 3.3:
```{r ridge}
X <- as.matrix(std_train_data[, c("lcavol", "lweight", "age", "lbph", "svi", "lcp", "gleason", "pgg45")])
y <- std_train_data$lpsa

# Ridge regression (alpha = 0)
ridge_model <- glmnet(X, y, alpha = 0)

cv_ridge <- cv.glmnet(X, y, alpha = 0, nfolds = 10)
best_lambda_ridge <- cv_ridge$lambda.min
ridge_final <- glmnet(X, y, alpha = 0, lambda = best_lambda_ridge)
print(coef(ridge_final))
ridge_coef <- coef(ridge_final)

X_test <- as.matrix(std_test_data[, c("lcavol", "lweight", "age", "lbph", "svi", "lcp", "gleason", "pgg45")])
y_test <- std_test_data$lpsa
ridge_preds <- predict(ridge_final, newx = X_test)
ridge_mse <- mean((y_test - ridge_preds)^2)
print(paste("Ridge MSE:", ridge_mse))
```
Column 5 of table table 3.3
```{r lasso}
# Lasso regression (alpha = 1)
lasso_model <- glmnet(X, y, alpha = 1)

# Cross-validation for best lambda
cv_lasso <- cv.glmnet(X, y, alpha = 1, nfolds = 10)
best_lambda_lasso <- cv_lasso$lambda.min

# Fit Lasso with best lambda
lasso_final <- glmnet(X, y, alpha = 1, lambda = best_lambda_lasso)
print(coef(lasso_final))
lasso_coefs <- coef(lasso_final)

X_test <- as.matrix(std_test_data[, c("lcavol", "lweight", "age", "lbph", "svi", "lcp", "gleason", "pgg45")])
y_test <- std_test_data$lpsa

lasso_preds <- predict(lasso_final, newx = X_test)
lasso_mse <- mean((y_test - lasso_preds)^2)
print(paste("Lasso MSE:", lasso_mse))
```
For best sub set selection, we keep removing the the least significant variable until they are all significant:
```{r b linear model fit}
model <- lm(lpsa ~ lcavol + lweight + age + lbph + svi + lcp  + pgg45, data = std_train_data)
model_summary <- summary(model)
results <- data.frame(
  Coefficient = model_summary$coefficients[, "Estimate"],
  Std_Error = model_summary$coefficients[, "Std. Error"],
  Z_Score = model_summary$coefficients[, "t value"],
  P_Value = model_summary$coefficients[, "Pr(>|t|)"]
)

print(round(results,3), row.names = FALSE)
```


Now remove age:
```{r b linear model fit}
model <- lm(lpsa ~ lcavol + lweight + lbph + svi + lcp + pgg45, data = std_train_data)
model_summary <- summary(model)
results <- data.frame(
  Coefficient = model_summary$coefficients[, "Estimate"],
  Std_Error = model_summary$coefficients[, "Std. Error"],
  Z_Score = model_summary$coefficients[, "t value"],
  P_Value = model_summary$coefficients[, "Pr(>|t|)"]
)

print(round(results,3), row.names = FALSE)
```

Now remove lcp
```{r b linear model fit}
model <- lm(lpsa ~ lcavol + lweight + lbph + svi + pgg45, data = std_train_data)
model_summary <- summary(model)
results <- data.frame(
  Coefficient = model_summary$coefficients[, "Estimate"],
  Std_Error = model_summary$coefficients[, "Std. Error"],
  Z_Score = model_summary$coefficients[, "t value"],
  P_Value = model_summary$coefficients[, "Pr(>|t|)"]
)

print(round(results,3), row.names = FALSE)
```

Now remove pgg45
```{r b linear model fit}
model <- lm(lpsa ~ lcavol + lweight + lbph + svi, data = std_train_data)
model_summary <- summary(model)
results <- data.frame(
  Coefficient = model_summary$coefficients[, "Estimate"],
  Std_Error = model_summary$coefficients[, "Std. Error"],
  Z_Score = model_summary$coefficients[, "t value"],
  P_Value = model_summary$coefficients[, "Pr(>|t|)"]
)

print(round(results,3), row.names = FALSE)
```
Now remove lbph
```{r b linear model fit}
model <- lm(lpsa ~ lcavol + lweight + svi, data = std_train_data)
model_summary <- summary(model)
results <- data.frame(
  Coefficient = model_summary$coefficients[, "Estimate"],
  Std_Error = model_summary$coefficients[, "Std. Error"],
  Z_Score = model_summary$coefficients[, "t value"],
  P_Value = model_summary$coefficients[, "Pr(>|t|)"]
)

print(round(results,3), row.names = FALSE)
```

Now remove svi (column 3 of table 3.3)
```{r b linear model fit}
model_best_subset <- lm(lpsa ~ lcavol + lweight, data = std_train_data)
model_summary_best_subset <- summary(model_best_subset)
results <- data.frame(
  Coefficient = model_summary_best_subset$coefficients[, "Estimate"],
  Std_Error = model_summary_best_subset$coefficients[, "Std. Error"],
  Z_Score = model_summary_best_subset$coefficients[, "t value"],
  P_Value = model_summary_best_subset$coefficients[, "Pr(>|t|)"]
)

print(round(results,3), row.names = FALSE)

y_pred_best_subset <- predict(model_best_subset, newdata = std_test_data)
y_test <- std_test_data$lpsa
prediction_errors <- (y_test - y_pred_best_subset)^2
mean_prediction_error_best_subset <- mean(prediction_errors)
print(paste("Best Subset MSE:", mean_prediction_error_best_subset))
```


Column 2 was obtain earlier when we fitted the linear regression.







This is now the best subset.
The prediction errors of the models:
```{r predict}
# best subset
y_pred_best_subset <- predict(model_best_subset, newdata = std_test_data)
y_test <- std_test_data$lpsa
prediction_errors <- (y_test - y_pred_best_subset)^2
mean_prediction_error_best_subset <- mean(prediction_errors)
print(paste("Best Subset MSE:", mean_prediction_error_best_subset))
print(paste("LS MSE:", mean_prediction_error))

```

















