---
title: "Homework 1"
author: "Jake Mc Leroth"
date: "2025-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r install libraries}
# uncomment if not installed already
# install.packages("MASS")
# install.packages("class")
# install.packages("ggplot2")
# install.packages("dplyr")
# install.packages("mvtnorm")
```


```{r libraries}
library(MASS)
library(class)
library(ggplot2)
library(dplyr)
library(mvtnorm)
set.seed(1)
```

Question 1:
a) The data generating process used for figure 2.2 starts by simulating from a bivariate normal distribution. We generate 10 values from N((1,0)',I), where I is a diagonal matrix of ones. Thus we will have 10 vectors each of length 2. These now represent the mean values for the blue dots. For the orange dots, we get the mean values using the same method but we generate values form the N((0,1)', I) bivariate normal distribution instead. No we have 10 mean vectors for blue and 10 mean vector for orange. 

Now, we want to generate 100 blue's and 100 orange's. Starting with the blues, we pick one of the mean vectors randomly (each one has a 0.1 chance of being picked), say blue_mean_k. Then, we generate a value from N(blue_mean_k, I/5). We repeat this to get 100 observations. The methedology is exactly the same for the oranges. This is shown in the next code snippet:
```{r simulate blue and orange points}
blue_means <- mvrnorm(10, c(1,0), matrix(c(1,0,0,1), nrow=2))
orange_means <- mvrnorm(10, c(0,1), matrix(c(1,0,0,1), nrow=2))

blues <- matrix(NA, nrow = 100, ncol = 2)
oranges <- matrix(NA, nrow = 100, ncol = 2)

for (i in 1:100) {
  blues[i,] <- mvrnorm(1, mu = blue_means[sample(1:10,1), ], matrix(c(1/5,0,0,1/5), nrow=2))
  oranges[i,] <- mvrnorm(1, mu = orange_means[sample(1:10,1), ], matrix(c(1/5,0,0,1/5), nrow=2))
}

```

We plot the points with the decisiono boundary:
```{r plot Figure 2.2}
# Combine points to fit linear regression
combined_data <- data.frame(
  x1 = c(oranges[,1], blues[,1]),
  x2 = c(oranges[,2], blues[,2]),
  class = factor(rep(c(1, 0), each = 100))  # 0=blue, 1=orange
)

lin_model <- lm(as.numeric(as.character(class)) ~ x1 + x2, data = combined_data)

# Calculate decision boundary from 0.5 = B_0 + B_1x1 + B_2x2 => (0.5 - B_0)/B_2 - B_1x1/B_2 (essentailly solving for x2)
intercept <- (0.5 - coef(lin_model)[1]) / coef(lin_model)[3]
slope <- -coef(lin_model)[2] / coef(lin_model)[3]

# Plot points with decision boundary
ggplot(combined_data, aes(x = x1, y = x2, color = class)) +
  geom_point(size = 3, alpha = 0.6) +
  scale_color_manual(values = c("blue", "orange"), 
                     labels = c("Blue (0)", "Orange (1)")) +
  geom_abline(intercept = intercept, 
              slope = slope, 
              color = "black",
              linewidth = 1.2,
              linetype = "solid") +
  labs(title = "Decision Boundary from Linear Regression (Figure 2.1 Replicate)") +
  theme_minimal() +
  theme(legend.position = "top")
```


b) Here we fit a model for the k-nearest neighbours algorithm, for k=15, and plot its decision boundary:
```{r Nearest Neighbour}

X <- combined_data[, c("x1", "x2")]
y <- combined_data$class

# Create grid for decision boundary visualization
grid_resolution <- 0.05
x1_range <- seq(min(combined_data$x1), max(combined_data$x1), by = grid_resolution)
x2_range <- seq(min(combined_data$x2), max(combined_data$x2), by = grid_resolution)
grid <- expand.grid(x1 = x1_range, x2 = x2_range)

# Fit kNN (k=15)
knn_pred <- knn(train = X, test = grid, cl = y, k = 15, prob = TRUE)

# Add predictions to grid
grid$class <- knn_pred
grid$prob <- attr(knn_pred, "prob")
grid$prob[grid$class == 0] <- 1 - grid$prob[grid$class == 0]  # Prob for class 0


ggplot() +
  # Grid
  geom_tile(data = grid, 
            aes(x = x1, y = x2, fill = class, alpha = prob),
            show.legend = FALSE) +
  scale_fill_manual(values = c("blue", "orange")) +
  scale_alpha_continuous(range = c(0.1, 0.3)) +
  
  # Original Blue and Orange points
  geom_point(data = combined_data, 
             aes(x = x1, y = x2, color = class),
             size = 3, alpha = 0.7) +
  scale_color_manual(values = c("blue", "orange")) +
  
  # kNN boundary (where prob = 0.5)
  geom_contour(data = grid, 
               aes(x = x1, y = x2, z = as.numeric(class)),
               bins = 1, color = "red", linewidth = 1) +
  
  labs(title = "kNN (k=15) Decision Boundary vs Linear Regression (Figure 2.2 Replicate)",
       subtitle = "Red line: kNN boundary | Dashed black: Linear regression boundary",
       x = "x1", y = "x2") +
  theme_minimal()
```

c) Now we use the mean values we simulated earlier to get the blue and oranges mixture distributions, in which we use to plot the optimal bayes decision boundary (i.e. p=0.5)
```{r}
# Mixture Densities
p_blue <- function(x) {
  mean(apply(blue_means, 1, function(mu) {
    dmvnorm(x, mean = mu, sigma = matrix(c(1/5, 0, 0, 1/5), nrow = 2))
  }))
}

p_orange <- function(x) {
  mean(apply(orange_means, 1, function(mu) {
    dmvnorm(x, mean = mu, sigma = matrix(c(1/5, 0, 0, 1/5), nrow = 2))
  }))
}

# Create grid
grid <- expand.grid(
  x1 = seq(min(c(blues[,1], oranges[,1])) - 1, 
           max(c(blues[,1], oranges[,1])) + 1, 
           length.out = 200),
  x2 = seq(min(c(blues[,2], oranges[,2])) - 1, 
           max(c(blues[,2], oranges[,2])) + 1, 
           length.out = 200)
)

# Calculate Bayes probabilities
grid$bayes_prob <- apply(grid, 1, function(x) {
  px0 <- p_blue(x)
  px1 <- p_orange(x)
  px1 / (px0 + px1)  # P(Y=1|X=x)
})

# Extract 0.5 probability contour
bayes_boundary <- contourLines(
  x = unique(grid$x1),
  y = unique(grid$x2),
  z = matrix(grid$bayes_prob, nrow = length(unique(grid$x1))),
  levels = 0.5
)

# Convert to plottable format
boundary_df <- data.frame(
  x1 = bayes_boundary[[1]]$x,
  x2 = bayes_boundary[[1]]$y
)

# Plot with Bayes boundary
ggplot() +
  geom_point(data = combined_data, aes(x1, x2, color = class), alpha = 0.6) +
  geom_path(data = boundary_df, aes(x1, x2), 
            color = "black", linewidth = 1.5) +
  scale_color_manual(values = c("blue", "orange")) +
  labs(title = "Bayes Decision Boundary (Figure 2.5 Replicate)") +
  theme_minimal()
```