---
title: "Homework 4"
author: "Jake Mc Leroth"
date: "2025-04-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries install, include=FALSE}
# install.packages("foreign")
# install.packages("mldr")
# install.packages("utiml")
# install.packages("corrplot")
# install.packages("randomForest")
# install.packages("e1071")
# install.packages("xgboost")
```

```{r libraries, results='hide', message=FALSE}
library(foreign)
library(mldr)
library(utiml)
library(corrplot)
library(randomForest)
library(e1071)
library(xgboost)
library(ggplot2)
library(reshape2)
```

#### In this assignment, the Scene data set is analysed. This is a multilabel data set from image data, with each image having binary labels for Beach, Sunset, Fall Folliage, Field, Mountain and Urban. The data set was first introduced by Boutell et al. (2004). The features for the data were obtained as follows:The RGB colours are converted to the Luv space. Each image was divided into a 7x7 grid of 49 squares, of which each has 3 bands corresponding to a low resolution image and texture features. The mean and variance is calculated for each bang, resulting in 49 * 3 * 2 = 294 features. The goal is to use these features to classify wether each image has a Beach, a Sunset, Fall Folliage, a Field, a Mountain or Urban.

```{r data}
# Import data
data <- foreign::read.arff("/Users/jakemcleroth/Desktop/University/Masters/Modules/Semester 1/SLT/homework/hw_4/scene_dat")
label_indices <- 295:300
data[label_indices] <- lapply(data[label_indices], function(x) as.numeric(as.character(x)))

# Create an mldr object
data_mldr <-  mldr_from_dataframe(data, labelIndices = label_indices, name = "scene")
```


## Exploratory Data Analysis
#### An exploratory data analysis is done on the data set:
```{r}
# Missing values
sum(colSums(is.na(data)))

# Number of columns
data_mldr$measures$num.attributes
# Number of observations
data_mldr$measures$num.instances
# Number of features
data_mldr$measures$num.inputs
# Number of labels
data_mldr$measures$num.labels
```
#### We see there are no missing values to deal with, and confirm there are 6 labels and 294 features. There are 2407 observation (images) in the data set. The labels are looked at in more depth, first looking at the individual labels:

```{r Individual label plots, out.width='50%'}
# Bar plot for label counts
label_count <- data_mldr$labels$count
barplot(label_count,
        main = "Label Count",
        ylab = "Count",
        las = 2, col = "yellow", names.arg = c("Beach", "Sunset", "FallFoliage", "Field", "Mountain", "Urban"))

# Bar plot for label frequency
label_freq <- data_mldr$labels$freq
barplot(label_freq,
        main = "Label Frequencies",
        ylab = "Proportion",
        las = 2, col = "steelblue", names.arg = c("Beach", "Sunset", "FallFoliage", "Field", "Mountain", "Urban"))

# Bar plot for Imbalance Ratio
barplot(data_mldr$labels$IRLbl,
        main = "Imbalance Ratio per Label (IRLbl)",
        ylab = "IR",
        las = 2,
        col = "tomato",
        names.arg = c("Beach", "Sunset", "FallFoliage", "Field", "Mountain", "Urban"))

# Correlation heat map
label_corr <- cor(data[, label_indices])
corrplot(label_corr,
         method = "color", type = "upper",
         tl.col = "darkgrey", tl.srt = 45,
         addCoef.col = "darkgrey", number.cex = 0.7)
```

#### We see from the first three plots that the most common label is Mountain, and the least is Sunset. There is quite a high within-label imbalance, with Mountain having a frequency of 20%. The correlation heat map shows that all labels exhibit a small negative correlation. We now look at the between label behaviour:

```{r}
labelsets <- data_mldr$labelsets
labelset_counts <- sort(labelsets, decreasing = TRUE)

barplot(labelset_counts,
        main = "Labelset Frequencies (All)",
        ylab = "Count",
        las = 2,
        col = "purple",
        names.arg = c("Urban", "Mountain", "Beach", "Sunset", "Fall Foliage", "Field", "F + M", "B + M", "FF + F", "B + U", "FF + M", "F + U", "M + U", "FF + F + M", "B + F"), cex.names = 0.8) #  names(labelset_counts)

labels_per_instance <- rowSums(data[, label_indices])
hist(labels_per_instance,
     breaks = max(labels_per_instance),
     main = "Number of Labels per Observation",
     xlab = "Label Count", col = "darkgreen")
```
#### There are 15 unique labelsets, 3 of which are only observed once. Urban and Mountain are the most common labelset, with the rest of the individual features following behind. The max labels present in any labelset is only two, with by far the majority of them only having one. Some more specific metrics are looked at:

```{r Multilabel measures}
# Cardinality
data_mldr$measures$cardinality

# Density
data_mldr$measures$density

# Mean imbalance ratio 
data_mldr$measures$meanIR

# Scumble
data_mldr$measures$scumble

# Scumble CV
data_mldr$measures$scumble.cv

# TCS
data_mldr$measures$tcs
cor_matrix <- 
```

#### *Interpret. Now the features are looked at. Since there are 294 features, only a few summary statistics are looked at to get a better idea of the data, since a 294x294 correlation plot is not very interpretable.

```{r Features} 
# Assume feature_data is your numeric data frame
cor_matrix <- cor(data[,-(295:300)])

# Turn matrix into long format
cor_df <- melt(cor_matrix, varnames = c("Feature1", "Feature2"), value.name = "Correlation")

# Plot correlation heat map
ggplot(cor_df, aes(x = Feature1, y = Feature2, fill = Correlation)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name = "Correlation") +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),  # remove x-axis text
    axis.text.y = element_blank(),  # remove y-axis text
    axis.ticks = element_blank(),   # remove axis ticks
    panel.grid = element_blank(),    # optional: remove grid lines
    legend.position = "right"
  ) +
  labs(title = "Feature Correlation Heatmap", x = "", y = "")

```

To do: feature analysis 
* Get ranges, max & min, means, check if any features have outliers + interpret multilabel measures


```{r}
# train/test split
set.seed(123)
train_indices <- sort(sample(1:nrow(data), size = 1211, replace = FALSE), decreasing = F)
data_train <- data[train_indices,]
data_test <- data[-train_indices,]
data_train_mldr <- mldr_from_dataframe(data_train, labelIndices = label_indices, name = "scene train")
data_test_mldr <- mldr_from_dataframe(data_test, labelIndices = label_indices, name = "scene test")
```

## Model Fitting
#### Transformation Models: Binary relevance, classifier chains, label powerset, calibrated label ranking 
```{r Binary Relevance}
# Binary relevance with RF
br_fit_rf <- br(data_train_mldr, base.algorithm = "RF", cores = 6)
br_predict_rf <- predict(br_fit_rf, data_test_mldr)
results_br_rf <- multilabel_evaluate(data_test_mldr, br_predict_rf, labels = T)

# Binary relevance with SVM
br_fit_svm <- br(data_train_mldr, base.algorithm = "SVM", cores = 6)
br_predict_svm <- predict(br_fit_svm, data_test_mldr)
results_br_svm <- multilabel_evaluate(data_test_mldr, br_predict_svm, labels = T)

# Binary relevance with xgboost
br_fit_xg <- br(data_train_mldr, base.algorithm = "XGB", cores = 6)
br_predict_xg <- predict(br_fit_xg, data_test_mldr)
results_br_xg <- multilabel_evaluate(data_test_mldr, br_predict_xg, labels = T)

```

```{r Classifier Chains}
# Classifier Chains with RF
cc_fit_rf <- cc(data_train_mldr, base.algorithm = "RF", cores = 6)
cc_predict_rf <- predict(cc_fit_rf, data_test_mldr)
results_cc_rf <- multilabel_evaluate(data_test_mldr, cc_predict_rf, labels = T)

# Classifier Chains with SVM
cc_fit_svm <- cc(data_train_mldr, base.algorithm = "SVM", cores = 6)
cc_predict_svm <- predict(cc_fit_svm, data_test_mldr)
results_cc_svm <- multilabel_evaluate(data_test_mldr, cc_predict_svm, labels = T)

# Classifier Chains with xgboost
cc_fit_xg <- cc(data_train_mldr, base.algorithm = "XGB", cores = 6)
cc_predict_xg <- predict(cc_fit_xg, data_test_mldr)
results_cc_xg <- multilabel_evaluate(data_test_mldr, cc_predict_xg, labels = T)
```


```{r Label powerset}
# Label powerset with RF
lp_fit_rf <- lp(data_train_mldr, base.algorithm = "RF", cores = 6)
lp_predict_rf <- predict(lp_fit_rf, data_test_mldr)
results_lp_rf <- multilabel_evaluate(data_test_mldr, lp_predict_rf, labels = T)

# Label powerset with SVM
lp_fit_svm <- lp(data_train_mldr, base.algorithm = "SVM", cores = 6)
lp_predict_svm <- predict(lp_fit_svm, data_test_mldr)
results_lp_svm <- multilabel_evaluate(data_test_mldr, lp_predict_svm, labels = T)

# Label powerset with xgboost
lp_fit_xg <- lp(data_train_mldr, base.algorithm = "XGB", cores = 6)
lp_predict_xg <- predict(lp_fit_xg, data_test_mldr)
results_lp_xg <- multilabel_evaluate(data_test_mldr, lp_predict_xg, labels = T)

```

```{r}
# multilabel_evaluate(data_test_mldr, lp_predict_xg, "bipartition")
# mldr_evaluate(data_test_mldr, lp_predict_xg)
# multilabel_confusion_matrix(data_test_mldr, lp_predict_xg)
# multilabel_evaluate(multilabel_confusion_matrix(data_test_mldr, lp_predict_xg))
# multilabel_evaluate(data_test_mldr, lp_predict_xg, "label-based", labels = T)
# multilabel_evaluate(data_test_mldr, fixed_threshold(lp_predict_xg, threshold = 0.5), "example-based", labels = T)
# mldr_evaluate(data_test_mldr, lp_predict_xg)
```

```{r Calibrated label ranking}

# Calibrated label ranking with RF
clr_fit_rf <- clr(data_train_mldr, base.algorithm = "RF", cores = 6)
clr_predict_rf <- predict(clr_fit_rf, data_test_mldr)
results_clr_rf <- multilabel_evaluate(data_test_mldr, clr_predict_rf, labels = T)

# Calibrated label ranking with SVM
clr_fit_svm <- clr(data_train_mldr, base.algorithm = "SVM", cores = 6)
clr_predict_svm <- predict(clr_fit_svm, data_test_mldr)
results_clr_svm <- multilabel_evaluate(data_test_mldr, clr_predict_svm, labels = T)

# Calibrated label ranking with xgboost
clr_fit_xg <- clr(data_train_mldr, base.algorithm = "XGB", cores = 6)
clr_predict_xg <- predict(clr_fit_xg, data_test_mldr)
results_clr_xg <- multilabel_evaluate(data_test_mldr, clr_predict_xg, labels = T)

```


#### Adaptation Models: KNN, Decision trees
```{r KNN}
# Multi-label knn
mlknn_fit <- mlknn(data_train_mldr)
mlknn_predict <- predict(mlknn_fit, data_test_mldr)
results_mlknn <- multilabel_evaluate(data_test_mldr, mlknn_predict, labels = T)
```

#### Ensemble Methods Ensemble br, Ensemble cc
```{r Ensemble binary relevance}
# Ensemble binary relevance with RF
ebr_fit_rf <- ebr(data_train_mldr, base.algorithm = "RF", cores = 6)
ebr_predict_rf <- predict(ebr_fit_rf, data_test_mldr)
results_ebr_rf <- multilabel_evaluate(data_test_mldr, ebr_predict_rf, labels = T)

# Ensemble binary relevance with SVM
ebr_fit_svm <- ebr(data_train_mldr, base.algorithm = "SVM", cores = 6)
ebr_predict_svm <- predict(ebr_fit_svm, data_test_mldr)
results_ebr_svm <- multilabel_evaluate(data_test_mldr, ebr_predict_svm, labels = T)

# Ensemble binary relevance with xgboost
ebr_fit_xg <- ebr(data_train_mldr, base.algorithm = "XGB", cores = 6)
ebr_predict_xg <- predict(ebr_fit_xg, data_test_mldr)
results_ebr_xg <- multilabel_evaluate(data_test_mldr, ebr_predict_xg, labels = T)

```


```{r Ensemble classifier chains}
# Ensemble classifier Chains with RF
ecc_fit_rf <- ecc(data_train_mldr, base.algorithm = "RF", cores = 6)
ecc_predict_rf <- predict(ecc_fit_rf, data_test_mldr)
results_ecc_rf <- multilabel_evaluate(data_test_mldr, ecc_predict_rf, labels = T)

# Classifier Chains with SVM
ecc_fit_svm <- ecc(data_train_mldr, base.algorithm = "SVM", cores = 6)
ecc_predict_svm <- predict(ecc_fit_svm, data_test_mldr)
results_ecc_svm <- multilabel_evaluate(data_test_mldr, ecc_predict_svm, labels = T)

# Classifier Chains with xgboost
ecc_fit_xg <- ecc(data_train_mldr, base.algorithm = "XGB", cores = 6)
ecc_predict_xg <- predict(ecc_fit_xg, data_test_mldr)
results_ecc_xg <- multilabel_evaluate(data_test_mldr, ecc_predict_xg, labels = T)
```



## References
#### Boutell, M.R., Luo, J., Shen, X. and Brown, C.M. (2004). Learning multi-label scene classification. Pattern recognition, vol. 37, no. 9, pp. 1757–1771.






























































































































































